---
title: "homework2"
output: pdf_document
date: "2023-03-25"
---
# Load dependencies
```{r load_hidden, echo=FALSE, warning=FALSE, results="hide"}
suppressPackageStartupMessages({
  library(devtools)
  library(Biobase)
  library(sva)
  library(bladderbatch)
  library(broom)
  library(tidyverse)
  library(data.table)
  library(RColorBrewer)
  library(gplots)
  library(broom)
  library(ggplot2)
  library(tidyverse)
})
```
```{r load}
  library(devtools)
  library(Biobase)
  library(sva)
  library(bladderbatch)
  library(broom)
  library(tidyverse)
  library(data.table)
  library(RColorBrewer)
  library(gplots)
  library(broom)
  library(ggplot2)
  library(tidyverse)
```

Loading the data and preprocessing, as it was done in the laboratory.

```{r}
library(bladderbatch)
data(bladderdata)

# sample info
pheno = pData(bladderEset)
# expression data
edata = exprs(bladderEset)

sumna <- apply(edata, 1, function(x) sum(is.na(x)))
row.variances <- apply(edata, 1, function(x) var(x))
edata <- edata[row.variances < 6,]
edata.log <- log2(edata)
```

> *Homework Problem 1:*
> Create a table to show the batch effects (refer to Figure 1 in Gilad and Mizrahi-Man, 2015). There are 5 batches (`pheno$batch`); how are biological variables and other variables related to study design are distributed among those 5 batches? Explain what could be a problem. Prepare this into a PDF file.

```{r}
library(gridExtra)
library(grid)

pdf('Cupriak_problem1.pdf')

df1 <- data.frame('batch1'=pheno[pheno$batch == 1,"outcome"][!duplicated(pheno[pheno$batch == 1,"outcome"])],
           'batch2'=pheno[pheno$batch == 2,"outcome"][!duplicated(pheno[pheno$batch == 2,"outcome"])],
           'batch3'=pheno[pheno$batch == 3,"outcome"][!duplicated(pheno[pheno$batch == 3,"outcome"])],
           'batch4'=pheno[pheno$batch == 4,"outcome"][!duplicated(pheno[pheno$batch == 4,"outcome"])],
           'batch5'=pheno[pheno$batch == 5,"outcome"][!duplicated(pheno[pheno$batch == 5,"outcome"])]
           )
levels(df1$batch1) <- c(levels(df1$batch1), '')
levels(df1$batch3) <- c(levels(df1$batch3), '')
levels(df1$batch4) <- c(levels(df1$batch4), '')
df1[2, 'batch1'] <- ''
df1[3, 'batch1'] <- ''
df1[2, 'batch3'] <- ''
df1[3, 'batch3'] <- ''
df1[2, 'batch4'] <- ''
df1[3, 'batch4'] <- ''


df2 <- data.frame('batch1'=pheno[pheno$batch == 1,"cancer"][!duplicated(pheno[pheno$batch == 1,"cancer"])],
           'batch2'=pheno[pheno$batch == 2,"cancer"][!duplicated(pheno[pheno$batch == 2,"cancer"])],
           'batch3'=pheno[pheno$batch == 3,"cancer"][!duplicated(pheno[pheno$batch == 3,"cancer"])],
           'batch4'=pheno[pheno$batch == 4,"cancer"][!duplicated(pheno[pheno$batch == 4,"cancer"])],
           'batch5'=pheno[pheno$batch == 5,"cancer"][!duplicated(pheno[pheno$batch == 5,"cancer"])]
           )
levels(df2$batch1) <- c(levels(df2$batch1), '')
levels(df2$batch2) <- c(levels(df2$batch2), '')
levels(df2$batch3) <- c(levels(df2$batch3), '')
levels(df2$batch4) <- c(levels(df2$batch4), '')
levels(df2$batch4) <- c(levels(df2$batch4), '')

df2[2, 'batch1'] <- ''
df2[2, 'batch3'] <- ''
df2[2, 'batch4'] <- ''

grid.table(df1)
plot.new()
grid.table(df2)
dev.off()
```

> *Homework Problem 2:*
> Make heatmaps, BEFORE and AFTER cleaning the data using ComBat, where columns are arranged according to the study design. You must sort the columns such that 5 batches are shown. Cluster the rows, but do not cluster the columns (samples) when drawing a heatmap. The general idea is that you want to see if the Combat-cleaned data are any improvement in the general patterns.

```{r}
library(sva)

# order the columns according to batch number
pheno.ordered <- pheno[order(pheno$batch),]
colnames.ordered <- row.names(pheno.ordered)
edata.ordered <- edata[,colnames.ordered]
batch.ordered <- pheno.ordered$batch

# clean edata using ComBat
combat_edata <- ComBat(dat=edata, batch=pheno$batch, mod=model.matrix(~1, data=pheno), par.prior=TRUE, prior.plots=TRUE)
combat_edata.oredered <- combat_edata[,colnames.ordered]
```

```{r}
## compare heatmaps before vs. after
library(gplots)
library(RColorBrewer)
my_palette <- colorRampPalette(c("blue", "white", "darkred"))(n = 299)

pdf("Cupriak_problem2.pdf",height=10,width=10)

colnames <- paste('batch', as.character(batch.ordered), ":", sep="")
colnames <- paste(colnames, colnames.ordered, sep=" ")

heatmap.2(edata.ordered,
          main = "Bladder Cancer Data before ComBat cleaning", # heat map title
          notecol="black",      # change font color of cell labels to black
          density.info="none",  # turns off density plot inside color legend
          trace="none",         # turns off trace lines inside the heat map
          margins =c(12,9),     # widens margins around plot
          col=my_palette,       # use on color palette defined earlier 
          dendrogram="none",
          scale = "row",
          Rowv = TRUE,
          Colv = FALSE,
          labCol = colnames)

heatmap.2(combat_edata.oredered,
          main = "Bladder Cancer Data after ComBat cleaning", # heat map title
          notecol="black",      # change font color of cell labels to black
          density.info="none",  # turns off density plot inside color legend
          trace="none",         # turns off trace lines inside the heat map
          margins =c(12,9),     # widens margins around plot
          col=my_palette,       # use on color palette defined earlier 
          dendrogram="none",
          scale = "row",
          Rowv = TRUE,
          Colv = FALSE,
          labCol = colnames)
dev.off()
```

> *Homework Problem 3:*
> Make heatmaps of Pearson correlations statistics of samples. For example, see Figure 2 and 3 freom Gilad and Mizrahi-Man (2015) F1000Research: \url{https://f1000research.com/articles/4-121}.
> First, compute the correlation statistics among columns. Second, create a heatmap using heatmap.2(). Make sure to create or add labels for samples (cancer vs. normal; batch numbers; others)

```{r}
# define my color palettes
color1 <- colorRampPalette(brewer.pal(8, "Spectral"))(25)
color2 <- colorRampPalette(brewer.pal(8, "PRGn"))(25)
color3 <- colorRampPalette(brewer.pal(8, "PuOr"))(25)

# compute the correlation statistics among columns
correlation <- cor(edata, method = "pearson")

colnames <- row.names(pheno)

pdf("Cupriak_problem3.pdf",height=10,width=10)
heatmap.2(correlation,
          col = color1,
          main = "Pearson correlation labeled cancer or normal", 
          notecol="black",      
          density.info="none",  
          trace="none",         
          margins =c(12,9), 
          dendrogram="none",
          labRow = paste(colnames, ":", pheno$cancer, sep = ""),
          labCol = paste(colnames, ":", pheno$cancer, sep = ""),
          cexRow=0.4,
          cexCol=0.5)

heatmap.2(correlation, 
          col = color2, 
          main = "Pearson correlation labeled by batch",
          notecol="black",      
          density.info="none",  
          trace="none",
          dendrogram = "none",
          margins =c(12,9),
          labRow = paste(colnames, ":batch", pheno$batch, sep=""),
          labCol = paste(colnames, ":batch", pheno$batch, sep=""),
          cexRow=0.4,
          cexCol=0.5)

heatmap.2(correlation, 
        col = color3, 
        main = "Pearson correlation labeled by outcome",
        notecol="black",      
        density.info="none",  
        trace="none",
        dendrogram = "none",
        margins =c(12,9),   
        labRow = paste(colnames, ":", pheno$outcome, sep = ""),
        labCol = paste(colnames, ":", pheno$outcome, sep = ""),
        cexRow=0.4,
        cexCol=0.5)

dev.off()
```

> *Homework Problem 4:*
> Apply two different Linear Models to the Bottomly et al. data. First, using a conventional approach, create a linear model with a genetic strain (biological variable) and an experimental number (technical variable) on **uncorrected** gene expression data. Second, create a linear model with a genetic strain (biological variables) on **corrected** gene expression data from ComBat. Make a scatter plots of coefficients and a histogram of p-values as done in this notebook. Make sure that you are pulling out the correct coefficients, not any or all coefficients. 

```{r load_bottomly_data}
con = url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bottomly_eset.RData")
load(file=con)
close(con)
edata_bottom <- as.matrix(exprs(bottomly.eset))
edata_bottom <- edata_bottom[rowMeans(edata_bottom) > 10, ]
edata_bottom <- log2(edata_bottom + 1)

pdata <- pData(bottomly.eset)

```

```{r}
# create a linear model with a genetic strain (biological variable) and an experimental number (technical variable) on uncorrected gene expression data from Bottomly.
model.uncorrected <- lm(t(edata_bottom) ~ as.factor(pdata$strain) + as.factor(pdata$experiment.number))

rownames(coefficients(model.uncorrected))

#tidy the data
model.uncorrected.tidy <- tidy(model.uncorrected)

# clean the data using ComBat
combat.edata_bottom = ComBat(dat=edata_bottom, batch=pdata$experiment.number, mod=model.matrix(~1, data=pdata), par.prior=TRUE, prior.plots=TRUE)

# create a linear model with a genetic strain (biological variable) and an experimental number (technical variable) on corrected gene expression data from Bottomly.
model.corrected <- lm(t(combat.edata_bottom) ~ as.factor(pdata$strain))

#tidy the data
model.corrected.tidy <- tidy(model.corrected)
```

```{r}
pdf("Cupriak_problem4.pdf", height = 10, width = 10)

# p-values related to genetic strain DBA/2J.

ggplot(model.uncorrected.tidy %>% filter(term == "as.factor(pdata$strain)DBA/2J")) + geom_histogram(aes(x=p.value), bins = 100, fill="darkorange") + ggtitle("P-values(strain DBA/2J) from linear model on UNCORRECTED data") + theme(plot.title = element_text(size=18))

ggplot(model.corrected.tidy %>% filter(term == "as.factor(pdata$strain)DBA/2J")) + geom_histogram(aes(x=p.value), bins = 100, fill="darkorange") + ggtitle("P-values (strain DBA/2J) from linear model on CORRECTED data") + theme(plot.title = element_text(size=18))

# scatter plots of coefficients

est_compare <- tibble(
  LinearModel = model.uncorrected.tidy %>% filter(term == "as.factor(pdata$strain)DBA/2J") %>% select("estimate") %>% unlist,
  ComBat = model.corrected.tidy %>% filter(term == "as.factor(pdata$strain)DBA/2J") %>% select("estimate") %>% unlist)

ggplot(est_compare, aes(x=LinearModel, y=ComBat)) +
     geom_point(col="darkgrey", alpha=.5, size=.5) + geom_abline(intercept=0, slope=1, col="darkred") + geom_smooth(method = "lm", se = TRUE)  + theme_bw() + ggtitle("Scatter plot of coefficients for genetic strain DBA/2J") + theme(plot.title = element_text(size=18))

dev.off()
```

## Problem 5
> *Homework Problem 5:*
> Apply ComBat and SVA to the Bottomly et al. data. 
> Make a scatter plots of coefficients and a histogram of p-values, comparing results based on ComBat and SVA. 
> Assume that the biological variables in Bottomly et al data is the genetic strains. 
> Make sure that you are pulling out the correct coefficients/pvalues, not any or all of them.

```{r}
mod = model.matrix(~as.factor(strain),data=pdata)
mod0 = model.matrix(~1, data=pdata)

sva_output = sva(edata_bottom, mod, mod0, n.sv=num.sv(edata_bottom,mod,method="leek"))
head(sva_output$sv)
summary(lm(sva_output$sv ~ pdata$experiment.number))
```

```{r}
# Add the surrogate variables to the model matrix
modsva = lm(t(edata_bottom) ~ as.factor(pdata$strain) + sva_output$sv)
modsva.tidy <- tidy(modsva)

pdf("Cupriak_problem5.pdf", height = 10, width = 10)

# scatter plot

est_compare <- tibble(
  ComBat = model.corrected.tidy %>% filter(term == "as.factor(pdata$strain)DBA/2J") %>% select("estimate") %>% unlist,
  SVA = modsva.tidy %>% filter(term == "as.factor(pdata$strain)DBA/2J") %>% select("estimate") %>% unlist)

ggplot(est_compare, aes(x=ComBat, y=SVA)) +
     geom_point(col="darkgrey", alpha=.5, size=.5) + geom_abline(intercept=0, slope=1, col="darkred") + geom_smooth(method = "lm", se = TRUE)  + ggtitle("Scatter plot of coefficients for genetic strain DBA/2J") + theme(plot.title = element_text(size=18))

# histogram of p-values

pvalues <- tibble(
  ComBat = model.corrected.tidy %>% filter(term == "as.factor(pdata$strain)DBA/2J") %>% select("p.value") %>% unlist,
  SVA = modsva.tidy %>% filter(term == "as.factor(pdata$strain)DBA/2J") %>% select("p.value") %>% unlist)

pvalues_gather <- gather(pvalues)
ggplot(pvalues_gather, aes(x=value)) + geom_histogram() + facet_wrap(~key)

dev.off()
```




